Supplemental Materials: \emph{Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours}

### 2015-03-26 Publications/2015-NeuroImage/04-methods.tex
Removing formulas from paper

% -*- root: 00-main.tex -*-
\section{Methods}
\label{sec:methods}
%

\subsection{Susceptibility-derived distortions}
\label{sec:distortions}
\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe the diffusion process within the brain in many different orientations,
  these sequences minimize the bandwidth of the \gls*{pe} direction.
This limitation originates a warping of the reconstructed image in regions affected by
  small deviations from the main magnetic field $B_0$.
One important source of these deviations ($\Delta B_0$) is the discontinuity of magnetic
  susceptibility across the imaged object.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bone of both hemispheres.

Distortion can be rendered as a registration problem where we look for an underlying deformation 
  field $U_s$, mapping from an undistorted \gls*{t1} image (the reference $R$) to the distorted
  \gls*{dmri} images (the moving space $M$), such that the structures identified in $R$ optimally
  align with the corresponding features in $M$:

  \begin{align}
  U_s\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

  \begin{equation}
  \vec{r}' = \vec{r} + \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \cdot \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel size along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution consisted on estimating $\Delta B_0$ with field mapping
  techniques \citep{andersson_modeling_2001}.
A second family of methods require an extra \gls*{epi} acquisition, but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient \gls*{pe}
  increments \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value.} of the \gls*{dmri} to an undistorted \gls*{t2} image of
   the same subject.
In the following, we will refer to this method as \gls*{t2b}.
Further developments \citep{irfanoglu_susceptibility_2011} extended the \gls*{t2b} 
  approach to the 3D problem.

\subsection{Contributions}
\label{sec:contributions}
%
While all the correction methods require an extra acquisition, \emph{regseg} is especially
  indicated in scanning protocols without specific data for correction,
  such as historical datasets.

Joint segmentation and distortion correction of \gls*{dmri} data is also interesting to
  improve the overall consistence of complex processing workflows.
The closest antecedent of \emph{regseg} is \emph{bbregister} \citep{greve_accurate_2009}.
Both methods use active contours to align surfaces to same-subject \gls*{dmri} images.
However, \emph{bbregister} is limited to affine transformations, and thus, it is not able to
  solve the susceptibility-derived distortions, generally nonlinear.
To enable registration of uncorrected data, \emph{bbregister} excludes from the cost-function
  those regions frequently warped.
Conversely, \emph{regseg} implements a free-deformation field, supported with B-Spline
  functions to estimate deformations.
A second diverging foundation is the active-contours framework.
In \emph{regseg}, active contours are implemented without edges \citep{chan_active_2001},
  enabling the tool to segment homogeneous regions within the brain.
On the other hand, \emph{bbregister} requires that the target features present intensity
  boundaries as it implements active contours with edges.
The tissue interfaces imaged with \gls*{dmri} are affected by partial volume effects that
  blur boundaries.
In practice, registration of \gls*{t1} and \gls*{dmri} data with \emph{bbregister}
  is limited to the linear match of the pial surface of the brain.

Finally, we present an instrumentation framework to evaluate correction and
  segmentation pipelines using realistic distortions as ground truth.
With the integration of \emph{regseg} and an in-house implementation of the 
  \gls*{t2b} correction in the framework, we compare the performance
  of both solutions.

@@

\subsection{Cost-function derivation}\label{sec:methods_map}
%
In a Bayesian framework, the mappings $U$ in \eqref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $R$.
Using the Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid R, \boldsymbol{\Gamma}) = \frac{P(R \mid U,\boldsymbol{\Gamma})\, P(U)}{P(R)},
  \label{eq:bayes_rule}
  \end{equation}
%
  where $P(R \mid U,\boldsymbol{\Gamma})$ is the data-likelihood, and
  $\boldsymbol{\Gamma} = \{\Gamma_{\text{in},\text{out}}\}$ is a set of surfaces 
  corresponding to the interfaces between pairs of competing regions $\Omega_\text{in}$
  and $\Omega_\text{out}$ drawn from the full set of regions $\{\Omega_l \mid l \in \mathbb{N}, l \leq L\}$.
As $\boldsymbol{\Gamma}$ are mapped by $U$, we simplify
  $P(U, \boldsymbol{\Gamma}) = P(U) \implies P(R \mid U,\boldsymbol{\Gamma}) = P(R \mid U)$.

The best estimate $\hat{U}$ then fulfills the maximum a posteriori criterium
  \citep{bishop_pattern_2006}, and aligns $\boldsymbol{\Gamma}$ into $R$.
First, we assume independence between pixels, and thus break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(R \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
%
  where $\vec{f}' = R(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the multispectral target
  volume.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$.
We then insert the normal distribution into \eqref{eq:bayes_rule} to obtain the full
  formulation of the \emph{data term} of our registration model:

 	\begin{align}
  P( R \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
%
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the reference image $R$.

@@

  \begin{align*}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \end{align*}
%
  imposing that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A} = (a_{ij})$ and $\mathbf{B} = (b_{ij})$.
Since the anisotropy is generally aligned with the imaging axes, $\mathbf{A}$ and $\mathbf{B}$
  can be simplified to diagonal matrices.
For convenience, we will set $\alpha_i = a_{ii}^2$ and $\beta_i = b_{ii}^2$, what can be reduced
  using the Hadamard square root notation\footnote{The Hadamard power of a matrix or a vector
  is the power of its elements $\mathbf{M}^{\circ p} = ({m_{ij}}^{p})$} as
  $\mathbf{A}=\boldsymbol{\alpha}^{\circ\frac12}\,\vec{I}_n$ and
  $\mathbf{B}=\boldsymbol{\beta}^{\circ\frac12}\,\vec{I}_n$.

Finally, the maximum a posteriori problem is adapted into a variational one where we look for
  the minimum of an energy functional, by applying a log-transform:

  \begin{align}
  E(R \mid U) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) \notag\\
  &= \const + \underset{l}{\sum} \int_{\Omega_l}
  \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} \left[ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} \right] \,d\vec{r},
  \label{eq:energy}
  \end{align}
%
  this expression is dual to the energy functional corresponding
  to a discrete \acrlong*{acwe} framework \citep{chan_active_2001}
  with the corresponding anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

\paragraph*{Deformation model}\label{sec:deformation_model}
Let us denote $\{\vec{v}_i\}_{i=1 \ldots N_c}$ the vertices of one or several prior
  surface(s).
In our application, these surfaces are triangularized meshes extracted using \emph{FreeSurfer}
  \citep{fischl_freesurfer_2012}.
The transform $\hat{U}$ \eqref{eq:transform} is supported by a dense deformation field
  $\vec{u} = u(\vec{r})$, such that:

  \begin{equation}
  \vec{v}_i' = \hat{U}\{\vec{v}_i\} = \vec{v}_i + u(\vec{v}_i) = \vec{v}_i + \vec{u}_i.
  \label{eq:nodes_tfm}
  \end{equation}

Since the nodes of the anatomical surfaces likely lay off-grid, it is required to
  derive $u(\vec{r})$ from a discrete set of parameters $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$.
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
  of degree 3: % ($\beta_3$).

  \begin{equation}
  u(\vec{r}) = \sum_k \psi_k(\vec{r}) \vec{u}_k.
  \label{eq:intp_kernel}
  \end{equation}

%Then, introducing \eqref{eq:intp_kernel} into \eqref{eq:nodes_tfm} and replacing
%  $\psi$ by the actual kernel function, the transformation writes:
%
%  \begin{equation}
%    \vec{v}_i' = \vec{v}_i + \sum_k \left[ \vec{u}_k \, \underset{d}{\prod}
%      \beta_3( (\vec{v}_i - \vec{r}_k) \cdot \hat{\mathbf{e}}_d ) \right],
%  \label{eq:transformation}
%  \end{equation}
%%
%  with $\hat{\mathbf{e}}_d$ being the unitary vector along axis $d$.

@@

  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(R \mid U)}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
%
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Now, we introduce \eqref{eq:energy} in \eqref{eq:general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}

We apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to \ref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

  \begin{align}
  \frac{\partial E_{data}(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \left\{
   \underset{\vec{x} \in \Omega_l}{\sum} \underset{l}{\sum} \mdist{f'}{l} \right\} \notag\\
  &= \underset{i}{\sum} w_i
   \left\langle \frac{\partial \vec{v}_i'}{\partial \vec{u}_k}, \bar{s}_i'\right\rangle,
  \label{eq:gradient_wshape}
  \end{align}
%
  in this case, the formulation has been adapted to the non-binary case, $\{l,m\}$
    being any pair of neighboring regions, and $\Gamma_{l,m}$ the contour separating
    them such that
    $\vec{x}' = \vec{v}' \in\Gamma_{l,m} \iff \vec{x}\in \partial\Omega_i \cap \partial\Omega_j$,
    $\bar{s}_i'$ is the speed vector projected on to the unit inward normal to the contour
    at $c_i'$ (described in \autoref{fig:method}), $w_i$ is the area contribution of vertex
    $c_i'$ to the the total area of the surface it belongs.

Finally, we can compute:

  \begin{align}
  \frac{\partial \vec{v}_i'}{\partial \vec{u}_k} &= \frac{\partial}{\partial \vec{u}_k}
  \left\{ \vec{v}_i + \sum_k \psi_k(\vec{v}_i) \vec{u}_k \right\}
  = \psi_k(\vec{v}_i)\, \hat{\vec{e}}
  \label{eq:basis_derivative}
  \end{align}
%
  where $\hat{\vec{e}}$ is the coordinates system's unit vector.
The vertex speeds $\bar{s}_i'$ obtained by computing the shape-gradients \eqref{eq:shape_gradients},
	are then projected to the deformation field in order to obtain the derivatives $\vec{g}_k$
	corresponding to $\vec{u}_k$:

  \begin{equation}
  \vec{g}_{i,k} = w_i \, \left\langle \frac{\partial}{\partial \vec{u}_k}{\vec{v}_i}', \bar{s}_i'\right\rangle
  = - w_i \left[ \mdist{f_i'}{l} - \mdist{f_i'}{m} \right] \psi_k(\vec{v}_i)\, \hat{\vec{n}}_i,
  \end{equation}
%
  then the full gradient evolution equation \eqref{eq:gradient_descent} yields:

  \begin{align}
  \frac{\partial E(\vec{u}_k)}{\partial \vec{u}_k} =
  &\underset{i}{\sum} \vec{g}_{i,k} +2\, \boldsymbol{\alpha} \vec{u}_k
  -2\, \boldsymbol{\beta} \Delta \vec{u}_k,
  \label{eq:gradient_final}
  \end{align}

Finally, we introduce a step size parameter $\delta$ to discretize the artificial time parameter
  in the iterative optimization.
Then, applying the Fourier transform as detailed in the \suppl{section S1}, the
  update equation is derived from \eqref{eq:gradient_final}:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\vec{u}_k^t / \delta - \sum_i \vec{g}_{i,k}\}}%
                  {\mathcal{F}\{(1/\delta+\boldsymbol{\alpha})I-\boldsymbol{\beta}\Delta\}} \right\}
  \label{eq:update_equation}
  \end{align}

>>>>

% -*- root: 00-main.tex -*-
\section{Methods}
\label{sec:methods}
%
Let $\gammaset = \{\Gamma_m: m \in \mathbb{N}, m \leq N_c\}$ be the set of $N_c$ surfaces
  extracted from the undistorted \gls*{t1} image (the reference space $R$).
Distortion can be rendered as a registration problem where we look for an underlying deformation 
  field $U_s$, mapping $R$ (with $\gammaset \subset R$) to the corresponding features in the distorted 
  \gls*{dmri} images (the moving space $M$):

  \begin{align}
  U_s\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
The displacements along the \gls*{pe} axis are related to the local field
  deviation $\Delta B_0(\vec{r})$ \citep{jezzard_correction_1995}:

  \begin{equation}
  u(\vec{r}) = \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \: \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel size along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.

% Joint segmentation and distortion correction of \gls*{dmri} data is also interesting to
%   improve the overall consistence of complex processing workflows.
% The closest antecedent of \emph{regseg} is \emph{bbregister} \citep{greve_accurate_2009}.
% Both methods use active contours to align surfaces to same-subject \gls*{dmri} images.
% However, \emph{bbregister} is limited to affine transformations, and thus, it is not able to
%   solve the susceptibility-derived distortions, generally nonlinear.
% To enable registration of uncorrected data, \emph{bbregister} excludes from the cost-function
%   those regions frequently warped.
% Conversely, \emph{regseg} implements a free-deformation field, supported with B-Spline
%   functions to estimate deformations.
% A second diverging foundation is the active-contours framework.
% In \emph{regseg}, active contours are implemented without edges \citep{chan_active_2001},
%   enabling the tool to segment homogeneous regions within the brain.
% On the other hand, \emph{bbregister} requires that the target features present intensity
%   boundaries as it implements active contours with edges.
% The tissue interfaces imaged with \gls*{dmri} are affected by partial volume effects that
%   blur boundaries.
% In practice, registration of \gls*{t1} and \gls*{dmri} data with \emph{bbregister}
%   is limited to the linear match of the pial surface of the brain.

@@

\subsection{Cost-function derivation}\label{sec:methods_map}
%
In a Bayesian framework for registration \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014},
  the mappings $U$ in \eqref{eq:transform} are
  evaluated based on their posterior probability given the observed data
  $M$.
Let $\omegaset = \{\Omega_l: l \in \mathbb{N}, l \leq L\}$ be the set of $L$ competing regions in
  which $M$ is partitioned by the projection of \gammaset, 
  the given surfaces serving as shape priors.
Using the Bayes' rule, the posterior likelihood is computed as:

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:bayes_rule}
  \end{equation}
%
  where $P(M \mid U,\omegaset)$ is the data-likelihood.
As $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.

The best estimate $\hat{U}$ then fulfills the maximum a posteriori criterium
  \citep{bishop_pattern_2006}, and aligns \gammaset into $M$.
First, we assume independence between pixels, and thus break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:bayes_aposteriori}
  \end{equation}
%
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:transform} in the multispectral target
  volume.
For convenience, and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions with parameters $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$.
We then insert the normal distribution into \eqref{eq:bayes_rule} to obtain the full
  formulation of the \emph{data term} of our registration model:

 	\begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} ( \vec{f} \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} ) \notag\\
  &= \underset{l}{\prod} \underset{\vec{r} \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:pdf}
  \end{align}
%
  using $\mdist{f}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}$ with respect
  to the descriptors of region $l$ as
  $\mdist{f}{l} = (\vec{f} - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f} - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the reference image $M$.

@@

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:thikonov}
  \end{align}
%
  imposing that the distortion and its gradient have zero
  mean and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Finally, the maximum a posteriori problem is adapted into a variational one where we look for
  the minimum of an energy functional, by applying a log-transform:

  \begin{align}
  E(M \mid U) &= -\log \underset{l}{\prod}
  \underset{\vec{r} \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) \notag\\
  &= \const + \underset{l}{\sum} \int_{\Omega_l}
  \mdist{f'}{l} \,d\vec{r} \, + \notag\\
  &+ \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}.
  \label{eq:energy}
  \end{align}
%

This expression is dual of the Mumford-Shah functional corresponding
  to a discrete \acrlong*{acwe} framework \citep{chan_active_2001}
  with the corresponding anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

\paragraph*{Deformation model}\label{sec:deformation_model}
Since the vertices of the surfaces $\{\vec{v}_i: \vec{v}_i \subset \gammaset \}_{i=1 \ldots V}$
  likely lay off-grid, it is required to derive $u(\vec{v}_i)$ from a discrete set of parameters
  $\{\vec{u}_k\}_{k=1 \ldots K}$.
Densification is achieved through a set of associated basis functions $\psi_k$.
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel
  of degree 3: % ($\beta_3$).

  \begin{equation}
  \vec{v}_i' = \vec{v}_i + u(\vec{v}_i) = \vec{v}_i + \sum_k \psi_k(\vec{r}) \: \vec{u}_k.
  \label{eq:nodes_tfm}
  \end{equation}

@@

  \begin{equation}
  \frac{\partial u(\vec{r},t)}{\partial t} \propto - \frac{\partial E(M \mid U)}{\partial \vec{u}_k},
  \label{eq:general_gradient_descent}
  \end{equation}
%
  with $t$ being an artificial time parameter of the contour
  evolution, and $\vec{u}_k$ the parameters supporting the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that anisotropy is aligned with the imaging axes to simplify
  \eqref{eq:energy} into expression \eqref{eq:app_energy} of \ref{app:reg_term}
  and compute the derivative \eqref{eq:general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \, + \notag\\
  &+ \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}
  \Big\}.
  \label{eq:gradient_descent}
  \end{align}
%
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The intermediate steps to get to \eqref{eq:gradient_descent} from \eqref{eq:final_gradient} are
  detailed in the \suppl{equations SM6-SM12}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \vec{g}_k  + \boldsymbol{\alpha} \cdot \vec{u}_k + \boldsymbol{\beta} \cdot (\Delta \vec{u}_k).
  \label{eq:final_gradient}
  \end{align}

Finally, to descend this gradient we establish a semi-implicit Euler scheme (see \suppl{section S1.3}),
  with a step size parameter $\delta$, that we solve it in the spectral domain as follows:

  \begin{align}
  \vec{u}_k^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{\delta^{-1} \, \vec{u}_k^t + \vec{g}_k\} }%
                  {\mathcal{F}\{(\delta^{-1} - \boldsymbol{\alpha})\, I-\boldsymbol{\beta}\Delta\}} \right\}
  \label{eq:update_equation}
  \end{align}

### 2015-03-26 Publications/2015-NeuroImage/08-appendix-00-shapepriors.tex
almost ready to revision round 2
update supplementals
improving to second round
Removing formulas from paper

% -*- root: 00-main.tex -*-
\renewcommand{\theequation}{A.\arabic{equation}}

@@

\subsection*{Instant shape priors}
\label{app:shape_priors}
Let $F(\vec{r})$ be an ``arbitrary'' function over the image domain
  $\Omega = \Omega_l \cup \Omega_m$ splitted in two regions $l$ and
  $m$, and $\Gamma_{l,m}$ a closed boundary between them
  as defined in \nameref{sec:methods_map}.
We now derive the domain integral w.r.t. $t$:

  \begin{align*}
  \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) d\vec{r} &=
  \int_\Omega \frac{\partial}{\partial t}F(\vec{r}) d\vec{r} \notag\\
  &- \int_{\Gamma_{l,m}} F(\vec{r}) \left\langle \frac{\partial \Gamma_{l,m} }{\partial t},
  N_{\Gamma_{l,m}}\right\rangle d\vec{r},
  \end{align*}

  where $\left\langle\frac{\partial\Gamma_{l,m}}{\partial t}, N_{\Gamma_{l,m}}\right\rangle$ is
  the projection of the boundary movement on the unit inward normal $N_{\Gamma_{l,m}}$.
Assuming that the region descriptors $\Theta_l$ vary slowly enough, we can consider
  that $\frac{\partial}{\partial t} F(\vec{r}) = 0$ and thus:

  \begin{equation*}
  \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) d\vec{r} =
  - \int_{\Gamma_{l,m}} F(\vec{r}) \left\langle \frac{\partial \Gamma_{l,m} }{\partial t},
  N_{\Gamma_{l,m}}\right\rangle d\vec{r}
  \end{equation*}

  In our formulation, $F(\vec{r})$ is the difference on
    sample likelihood between the inner ($\Omega_l$) and the outer ($\Omega_m$) regions.
  Therefore, we substitute:

    \begin{equation}
    \frac{\partial}{\partial t} \int_\Omega F(\vec{r}) d\vec{r} =
    - \int_{\Gamma_{l,m}} \left[ \mathcal{E}_{l}(\vec{r}) - \mathcal{E}_{m}(\vec{r}) \right]
    \left\langle \frac{\partial \Gamma_{l,m} }{\partial t},
  N_{\Gamma_{l,m}}\right\rangle d\vec{r},
  	\label{eq:shape_gradients}
    \end{equation}

    where $\mathcal{E}_l(\vec{r}) = \mathcal{D}^2_{l}(R(\vec{r}))$ for briefness.

>>>>

% -*- root: 00-main.tex -*-
\renewcommand{\theequation}{A.\arabic{equation}}
\renewcommand{\thesubsection}{Appendix \arabic{subsection}}

@@

\subsection{Simplifying the regularization term}\label{app:reg_term}
The exponentials of the Thikonov regularization prior \eqref{eq:thikonov} have the general form
  $\vec{v}^T \mathbf{M} \vec{v}$.
If $\mathbf{M}$ is a $n \times n$ diagonal matrix such that $\mathbf{M} = \vec{m} \, \mathbf{I}_n$, 
  then:
\begin{equation*}
\vec{v}^T \mathbf{M} \vec{v} = \vec{m} \cdot (\vec{v}^T \mathbf{I}_n \vec{v}) = \vec{m} \cdot \vec{v}^{\circ2},
\end{equation*}
%
  where we have introduced the Hadamard power notation\footnote{The Hadamard power of a matrix or a vector
  is the power of its elements $\mathbf{M}^{\circ p} = ({m_{ij}}^{p})$}.

Since the anisotropy is generally aligned with the imaging axes, 
  $\mathbf{A}$ and $\mathbf{B}$ of \eqref{eq:energy} can be simplified to diagonal matrices, such that
  $\mathbf{A}= \, \boldsymbol{\alpha}\,\vec{I}_n$ and
  $\mathbf{B}= \, \boldsymbol{\beta}\,\vec{I}_n$.
Substituting in the equation \eqref{eq:energy}, we get:
  \begin{align}
  E(M \mid U) &= \const + \underset{l}{\sum} \int_{\Omega_l}
  \mdist{f'}{l} \,d\vec{r} \notag\\
  &+ \int_{\Omega} \frac12 \left[ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2} + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} \right] \,d\vec{r}.
  \label{eq:app_energy}
  \end{align}
